---
title: "Tips annotated with growth temperature and coding genes/genome size"
date: "2024-08-21"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = FALSE,
    message = FALSE
)
```

```{r, message=FALSE}
library(bugphyzz)
library(purrr)
library(dplyr)
library(taxPPro)
library(ggplot2)
physNames <- c(
    "growth temperature", "coding genes", "genome size"
)
bp <- importBugphyzz(v = 0)
phys <- physiologies(physNames)
```

```{r}
bpSub <- bp[physNames] |> 
    map(~ mutate(.x, NCBI_ID = as.character(NCBI_ID))) |> 
    map(~ filter(.x, Evidence != "asr")) |> 
    map(as_tibble)
ps <- phys |> 
    map(~ mutate(.x, NCBI_ID = as.character(NCBI_ID))) |> 
    map(as_tibble)
```

```{r}
ltp <- ltp()
tipData <- as_tibble(ltp$tip_data)
```

There are 23,501 tips in the tree, including genera inserted with zero-length
branches:

```{r}
nrow(tipData)
```

Most of the tips correspond to species but there are some strains as well:

```{r}
tipData |> 
    count(Rank) |> 
    arrange(-n)
```

All tip labels in the LTP tree have names indicating an accession number,
e.g. *Salmonella\_enterica\_subsp.\_enterica-\_-AE006468--\_LT2\_T*.

I got he taxids from that accession number, e.g., *99287*.

These taxids were used for mapping the annotations in bugphyzz to the tips
in the LTP tree.

```{r}
tipData |> 
    filter(!grepl("g__", tip_label)) |> 
    select(tip_label, accession, taxid) |> 
    glimpse(width = 80)
```

If we look at the number of annotations per dataset before any propagation
is done,
we see that we only have less than 5,000 annotations for genome size and
coding genes. On the the other hand, we have more than 50,000 annotations
for growth temperature:

```{r}
map(phys, nrow)
```

When the annotations are mapped to the taxids in the tree, the percentage of
annotated tips is quite similar:

```{r}
map(phys, ~ {
    mean(tipData$taxid %in% .x$NCBI_ID) * 100
})
```

Even when there are so many annotations for growth temperature
(mostly strains from BacDive), they cannot be mapped to the LTP tree without
a taxid:

```{r}
phys |> 
    map(~ {
        .x |> 
            mutate(hasid = case_when(
                is.na(NCBI_ID) ~ "without id",
                is.integer(as.integer(NCBI_ID)) ~ "with id",
                TRUE ~ NA
            )) |> 
            count(Attribute_source, hasid)
    }) |> 
    bind_rows(.id = "physiology") |> 
    ggplot(aes(physiology, n)) +
    geom_col(aes(fill = Attribute_source)) +
    facet_wrap(~ hasid, nrow = 1) +
    theme_bw() +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1)
    )
```

So, after getting a consensus annotation for the species of the strains without 
taxid, now a good number of species can be mapped to the tree:

```{r}
map(bpSub, ~ {
    mean(tipData$taxid %in% .x$NCBI_ID) * 100
})

```

We can see that the number of taxa with ID now is much greater for
growth temperature:

```{r}
bpSub |> 
    map(~ {
        .x |> 
            mutate(hasid = case_when(
                is.na(NCBI_ID) ~ "without id",
                is.integer(as.integer(NCBI_ID)) ~ "with id",
                TRUE ~ NA
            )) |> 
            mutate(
                Attribute_source = ifelse(
                    is.na(Attribute_source), "tax (IBD)", Attribute_source
                )
            ) |> 
            count(Attribute_source, hasid)
    }) |> 
    bind_rows(.id = "physiology") |> 
    ggplot(aes(physiology, n)) +
    geom_col(aes(fill = Attribute_source)) +
    facet_wrap(~ hasid, nrow = 1) +
    theme_bw() +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1)
    )
```

Session information:

```{r}
sessioninfo::session_info()
```

